---
title: "Agglomerative Hierarchical Clustering for Anomaly Detection"
subtitle: Network Intrusion Detection via Zeek Log Analysis
author: "Sean P. Murphy"
date: "`r format(Sys.time(), '%d %B %Y')`"
format:
  revealjs: 
    theme: dark
editor: visual
bibliography: D214-BibTeX.bib
csl: apa-manual.csl
---


```{r Useful Libraries, include=FALSE}
# Data Manipulation
library(tidyverse)

# Visualizations
library(factoextra)
library(cluster)
library(ggplot2)
library(gridExtra)
library(kableExtra)

# Visualization Font
library(sysfonts)
library(showtextdb)
library(showtext)
font_add("LM Roman 10", "./lmroman10-regular.otf")
showtext_auto()
```

## Agenda
-   Introduction
-   Problem and Hypothesis 
-   Data Analysis Process    
-   Findings 
-   Limitations
-   Reccomendations 
-   Benefits


## Introduction
-   Graduate MSDA Student
-   Undergraduate Mathematics
-   Worked as Analyst for 13 Years
	- Linguistic Analysis
	- Intelligence Analysis
	- Signals Analysis
	
	
## Problem
- Network Intrusion Detection Systems (NIDS)
	- Signature Detection
	- Anomaly Detection
- PCAP vs Zeek
- Machine Learning
	- Agglomerative Hierarchical Clustering
[@mazarbhuiya2019], [@andrews2019]

## Hypothesis
- Assume Status Quo
	- No NIDS
	- No PCAP
	
$H_0$: A hierarchical clustering analysis of Zeek logs cannot reveal anomalous network behavior.

$H_A$: A hierarchical clustering analysis of Zeek logs can reveal anomalous network behavior.

[@sikos2020]


## Data Analysis Process
-   Data Collection and Preparation
-   Data Analysis


## Data Collection and Preparation
- Need Good Data
	- Zeek logs
	- Active Users
	- Hosted Services
	- Compromised
	
	
[@zhang2010]


## Data Collection
- U.S. Army Cyber Command (ARCYBER) 
  - Integrated Data Division (ID2)
- Approved for Release
- Extracted from the Gabriel Nimbus (GN)
- Transferred to a Private Server
	
[@brust2021]


## Data Preparation
- Raw Data
	- 23,747 Observations across 98 features
	- Seven Types of Logs
		- conn, dns, files, http, ssh, ssl, and weird 
- Subsetted Connectivity Logs
	- 10,726 Observations across 4 features
	- destination.packets, destination.port, source.packets, and source.port
	- 0% Sparse 
	
	[@thezeekproject]
	
	
## Data Preparation
- Scale Data
- Outliers Remain

[@mazarbhuiya2019]

## Analysis: Linkage Method

- Dissimilarity Matrix
- Linkage Methods
	- Single Linkage, Average Linkage, Complete Linkage, Ward's Method.

```{r AC Comparison, echo=FALSE}
ac.colname <- c("Average AC", "Single AC", "Complete AC", "Ward’s AC")
ac.data    <- c(0.9999928, 0.9999898, 0.9999935, 0.9999964)
ac         <- as.data.frame(t(ac.data))
  
knitr::kable(ac, caption = "Agglomerative Coefficient Comparison", col.names = ac.colname) %>%
   kable_styling(full_width = TRUE, latex_options = c("striped", "hold_position"))
```

## Analysis: Number of Clusters

```{r PRES-Silhouette, echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE}
df <- read_csv('./zeek.csv', col_names = TRUE, show_col_types = FALSE)

df.conn.scaled <- df %>% 
  filter(bdp.ingest.file.name == "conn.log") %>%
  select(destination.packets, destination.port, source.packets, source.port) %>%
  scale()

exsum.silhouette <- fviz_nbclust(df.conn.scaled, FUN = hcut, 
                      method = "silhouette", k.max = 10) +
                      labs(title = "Figure 1: Average Silhouette Width") +
                      theme_dark() +
                      theme(text = element_text(family = "LM Roman 10"))

exsum.silhouette
```

[@lengyel2019]


## Analysis: Cluster Membership

```{r PRES-Two-Cluster Membership Calc, echo=FALSE, cache=TRUE, fig.showtext=TRUE}
library(cluster)
df.conn.dist <- dist(df.conn.scaled, method = "euclidean")
hc.ward      <- agnes(df.conn.dist, method = "ward")
clust.2      <- cutree(hc.ward, k = 2)
```

```{r PRES-Two-Cluster Membership Viz, echo=FALSE, cache=TRUE, fig.showtext=TRUE}
fviz_cluster(list(data = df.conn.scaled, cluster = clust.2), geom = "point",
  font.family = "LM Roman 10") +
  labs(title = "Figure 2: Two-Cluster Observation Membership Plot", y=NULL, x=NULL) +
  annotate("text", x = -4, y = -35, label = "Anomalous Network Traffic") +
  annotate("text", x = -7.5, y = -4.5, label = "Normal Network Traffic") +
  theme_dark() +
  theme(text = element_text(family = "LM Roman 10"))
```

## Analysis: Cluster Assessment

- Cluster 1: 0.9686805
- Cluster 2: 0.9870862

[@rousseeuw1987]

## Findings: Summary

- Project Goals:
	- Isolate Anomalous Traffic as IoC

- Agglomerative Coefficients: > 0.9
- Average Silhouette Width: > 0.9
- Ward's Method
- Anomoly Density: 0.02% of Total Observations

## Interpretation

- Reject the Null Hypothesis

## Limitations

- Technique Limitation
	- Not equivalent to threat identification
	- Further Investigation Required

- Tool Limitation
	- Computational Cost
	- SageMaker instances increase > 15% 
	- Can't Usually Narrow Scope

[@amazonwebservices2022], [@zhang2019]


## AWS Costs Table

```{r AWS Cost Comparison, echo=FALSE}
aws.colname <- c("Base Specifications", "Upgraded Specifications")
aws.rowname <- c("vCPU", "Memory", "Clock Speed", "GPU", "NIC", "Storage", "GPU Memory", "Monthly Cost")
aws.before  <- c("4", "16 GB", "3.1 GHz", "N/A", "5 Gb", "EBS Only", "N/A", "$620.68")
aws.after   <- c("8", "32 GB", "2.5 GHz", "1", "25 Gb", "225 GB NVMe", "16 GB", "$713.18")
aws         <- as.data.frame(cbind(aws.before, aws.after), row.names = aws.rowname)

knitr::kable(aws, caption = "AWS SageMaker Cost/Specification Comparison", 
	col.names = aws.colname) %>%
  kable_styling(full_width = TRUE, latex_options = c("striped", "hold_position"))
```

## Proposed Actions

A summary of proposed actions:

- Implement Machine Learning

- Forward Results

## Benefits

Expected benefits of the study (be as specific and quantitative as possible):
 
- Increased Options

- Reduced Human Resources Required


## Increased Options Availability

- NIDS & Forensic Packet Analysis Strength
- Zeek
	- Practical
	- Opens Doors

[@sikos2020]


## Reduced Human Resource Requirements

- Provides Focus
	- Reduced Noise by 99.98%
- Save 1,787 Hours of Work
- Reduces MTTI by over a Month

[@mohit2022]
